---
title: "How to use caugi in a package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use caugi in a package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(caugi)
set.seed(42)
```

Now, let's see how you can use `caugi` in your own R package. We will work through an example to illustrate how you could approach this.

## The setup
Imagine that you want to build a causal discovery function that utilizes `caugi` for graph representation and manipulation. While seemingly not a very good idea, let's pretend your algorithmic idea is to measure the correlation between variables and then draw causal conclusions based on this^[Note that correlation does not imply causation!]. 

```{r first-iteration}
#' @title Correlation implies causation!
#'
#' @param df A `data.frame` with numeric columns
#'
#' @returns A `caugi` representing the causal graph that is totally true!
correlation_implies_causation <- function(df) {
  NULL # not developed yet!
}
```

Let's assume we have a named data frame:

```{r df-creation}
# create correlated data using MASS
df <- MASS::mvrnorm(
  n = 100,
  mu = c(0, 0, 0),
  Sigma = matrix(c(
    1, 0.8, 0.3,
    0.8, 1, 0.4,
    0.3, 0.4, 1
  ), nrow = 3)
) |> as.data.frame()
head(df)
```

## We know the nodes!

Now, we _know_ that the `caugi` should include all variables in the `df`. We don't know if the graph is a `DAG`, `PDAG`, or something else, so we will create a graph of the `UNKNOWN` class. We can begin with that, as a start:

```{r second-iteration}
#' @title Correlation implies causation!
#'
#' @param df A `data.frame` with numeric columns
#'
#' @returns A `caugi` representing the causal graph that is totally true!
correlation_implies_causation <- function(df) {
  cg <- caugi::caugi(nodes = names(df))
  return(NULL)
}
```

## Adding edges based on correlation

We can now compute the correlation matrix and add edges based on some arbitrary threshold:

```{r third-iteration}
#' @title Correlation implies causation!
#'
#' @param df A `data.frame` with numeric columns
#'
#' @returns A `caugi` representing the causal graph that is totally true!
correlation_implies_causation <- function(df) {
  cg <- caugi::caugi(nodes = names(df))
  cor_matrix <- cor(df)
  # Add edges for correlations above 0.5
  for (i in seq_len(ncol(cor_matrix))) {
    for (j in 1:i) {
      if (i != j && abs(cor_matrix[i, j]) > 0.5) {
        from <- names(df)[j]
        to <- names(df)[i]
        cg <- caugi::add_edges(cg, from = from, edge = "-->", to = to) # add edge to caugi
      }
    }
  }
  return(cg)
}
```

Now, when you call `correlation_implies_causation(df)`, it will return a `caugi` graph with edges based on the correlation threshold.

## Trying it out

Let's try it out!

```{r try-it-out}
cg <- correlation_implies_causation(df)
cg
```

The graph is ready to use. `caugi` uses lazy evaluation internally - the Rust graph session handles compilation and caching automatically when needed, so you don't need to manually build the graph.

## Class of the output

Now, you might want to specify the class of the output graph. Let's say that _if possible_ the output should be a DAG (which is advantageous for several reasons), but you don't want to enforce acyclicity in the algorithm, as that could sometimes cause your function to throw an error. You would rather have it return a graph in any case, but if it _is_ a DAG, then we return a DAG. 

```{r fifth-iteration}
#' @title Correlation implies causation!
#'
#' @param df A `data.frame` with numeric columns
#'
#' @returns A `caugi` representing the causal graph that is totally true!
correlation_implies_causation <- function(df) {
  cg <- caugi::caugi(nodes = names(df))
  cor_matrix <- cor(df)
  # Add edges for correlations above 0.5
  cg <- caugi::caugi(nodes = names(df))
  cor_matrix <- cor(df)
  # Add edges for correlations above 0.5
  for (i in seq_len(ncol(cor_matrix))) {
    for (j in 1:i) {
      if (i != j && abs(cor_matrix[i, j]) > 0.5) {
        from <- names(df)[j]
        to <- names(df)[i]
        cg <- caugi::add_edges(cg, from = from, edge = "-->", to = to) # add edge to caugi
      }
    }
  }
  if (caugi::is_dag(cg)) cg <- caugi::mutate_caugi(cg, class = "DAG")
  return(cg)
}
```

Now, when you call `correlation_implies_causation(df)`, it will return a `caugi` graph that is a DAG if possible, otherwise an `"UNKNOWN"` graph.

```{r try-it-out-2}
cg <- correlation_implies_causation(df)
cg
cg@graph_class
```

## That's it!

You have now successfully integrated `caugi` into your own R package function! Good luck and happy coding!
