---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

## Performance

This vignette discusses the performance characteristics of `caugi` including a comparison to the popular [`igraph` R package](https://igraph.org/r/) ([rigraph](https://github.com/igraph/rigraph)).
We focus on the practical trade-offs that arise from different core data structures and design choices.

The headline is: caugi frontloads computation.
That is, `caugi` spends more effort when constructing a graph and preparing indexes, so that queries are wicked fast. The high performance is also due to the Rust backend.

### Design choices

#### Compressed Sparse Row (CSR) representation

The core data structure in `caugi` is a Compressed Sparse Row (CSR) representation of the graph.
CSR stores for each vertex a contiguous slice of neighbor IDs with a pointer (offset) array that marks the start/end of each slice.
This format is memory efficient for sparse graphs. The `caugi` graph object also stores important query information in the object, leading to parent, child, and neighbor queries being done in $\mathcal{O}(1)$. This yields a larger memory footprint, but the trade-off is that queries are extremely fast.

#### Mutation and lazy building

The `caugi` graph objects are expensive to build. This is the performance downside of using `caugi`. For each time, we make a modification to a `caugi` graph object, we need to rebuild the graph completely, since the graph object is immutable by design. This has complexity $\mathcal{O}(|V| + |E|)$, where $V$ is the vertex set and $E$ is the edge set. 

However, the graph object will only be rebuild, when the user either calls `build()` directly or queries the graph. Therefore, you do not need to worry about wasting compute time by iteratively making changes to a `caugi` graph object, as the graph rebuilds lazily when queried. By doing this `caugi` graphs _feel_ mutable, but, in reality, they are not. 

By doing it this way, we ensure 
- that the graph object is always in a consistent state when queried, and
- that queries are as fast as possible,
while keeping the user experience smooth.

### Comparison

```{r setup}
library(caugi)
library(igraph)
library(microbenchmark)
```

We compare `caugi` to `igraph` on querying. 
We generate a random tree graph with 1000 nodes for the comparison.

```{r generate-graph}
ig <- igraph::make_tree(n = 1000)
cg <- caugi::as_caugi(ig, class = "DAG")
```
`
```{r benchmark-igraph-query}
mb_igraph <- microbenchmark(
  igraph_neighbors = igraph::neighbors(ig, 50),
  igraph_parents = igraph::neighbors(ig, 50, mode = "in"),
  igraph_children = igraph::neighbors(ig, 50, mode = "out"),
  times = 1000
)

mb_caugi <- microbenchmark(
  caugi_neighbors = caugi:::neighbors_of_ptr(cg@ptr, 50),
  caugi_parents = caugi:::parents_of_ptr(cg@ptr, 50),
  caugi_children = caugi:::children_of_ptr(cg@ptr, 50),
  times = 1000
)
```

```{r benchmark-results}
print(mb_igraph)
print(mb_caugi)
```

### Session info

```{r session-info}
sessionInfo()
```
