---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
```

```{r}
#| label: data
#| include: false

benchmarks <- readRDS("data/benchmarks.rds")
metadata <- jsonlite::fromJSON("data/metadata.json", simplifyVector = FALSE)
benchmarks$median_ms <- benchmarks$median_ns / 1e6
```

This article compares the performance of `caugi` against R packages
([`igraph`](https://r.igraph.org/), [`bnlearn`](https://www.bnlearn.com/),
[`dagitty`](https://dagitty.net/),
[`ggm`](https://cran.r-project.org/package=ggm),
[`pcalg`](https://pcalg.r-forge.r-project.org/)), the Python package
[`pgmpy`](https://pgmpy.org/), and the Java library
[Tetrad](https://www.cmu.edu/dietrich/philosophy/tetrad/). We focus on the
practical trade-offs that arise from different core data structures and design
choices.

The headline is: `caugi` frontloads computation. That is, `caugi` spends more
effort when constructing a graph and preparing indexes, so that queries are
wicked fast. The high performance is also due to the Rust backend.

A note on Tetrad: it is included as a reference point---the de facto golden
standard implementation in causal discovery---but the comparison is not strictly
apples-to-apples, since Tetrad runs on the JVM (compiled Java) while the R
packages and `caugi` run with R-level dispatch overhead on every call. Read the
Tetrad numbers as a "what's possible with compiled code" benchmark rather than a
head-to-head with the R-side competitors.

Benchmarks below are **precomputed**, with the last run on
`r metadata$generated` on host `r metadata$host$nodename`
(`r metadata$host$sysname`).

Versions:

- caugi `r metadata$versions$caugi`
- igraph `r metadata$versions$igraph`
- bnlearn `r metadata$versions$bnlearn`
- dagitty `r metadata$versions$dagitty`
- ggm `r metadata$versions$ggm`
- pcalg `r metadata$versions$pcalg`
- pgmpy `r metadata$versions$pgmpy`
- Tetrad `r metadata$versions$tetrad`.

## Design choices

### Compressed Sparse Row (CSR) Representation

The core data structure in `caugi` is a compressed sparse row (CSR)
representation of the graph. CSR representations store, for each vertex, a
contiguous slice of neighbor IDs with a pointer (offset) array that marks the
start and end of each slice. This format is memory-efficient for sparse graphs.
The `caugi` graph object also stores important query information in the object,
leading to parent, child, and neighbor queries to be done in $\mathcal{O}(1)$
time. This yields a larger memory footprint, but the trade-off is that queries
are extremely fast.

### Mutation and Lazy Building

The `caugi` graph objects are expensive to build. This is the performance
downside of using `caugi`. For each time we make a modification to a `caugi`
graph object, we need to rebuild the graph completely since the graph object is
immutable by design. This has complexity $\mathcal{O}(|V| + |E|)$, where $V$ is
the vertex set and $E$ is the edge set.

However, the graph object will only be rebuilt when the user either calls
`build()` directly or queries the graph. Therefore, you do not need to worry
about wasting compute time by iteratively making changes to a `caugi` graph
object, as the graph rebuilds lazily when queried. By doing this, `caugi` graphs
*feel* mutable, but, in reality, they are not.

By doing it this way, we ensure

- that the graph object is always in a consistent state when queried and
- that queries are as fast as possible,

while keeping the user experience smooth.

## Comparison

The comparison covers seven operations across a parameter grid of graph sizes
$n$ and edge densities. All packages run on the same generated DAG fixtures so
query results are directly comparable.

The grid covers $n \in \{100, 500, 1000\}$ at two density levels, parametrised
by the **average in-degree** $d$ (equivalently the average out-degree, or total
edges divided by number of nodes). We sample DAGs with per-edge probability
$p = 2d / (n - 1)$, which makes the expected in-degree exactly $d$ regardless of
$n$. The grid uses $d = 3$ for the sparser cells and $d = 6$ for the denser
ones, so a node has on average \~3 (or \~6) parents and \~3 (or \~6) children,
with edge counts that grow linearly with $n$. Each package runs against the same
generated graphs.

```{r}
#| label: helpers
#| include: false

library(ggplot2)

pkg_levels <- c(
  "caugi",
  "igraph",
  "bnlearn",
  "dagitty",
  "ggm",
  "pcalg",
  "pgmpy",
  "tetrad"
)
benchmarks$package <- factor(benchmarks$package, levels = pkg_levels)

plot_op <- function(op) {
  d <- benchmarks[benchmarks$operation == op, ]
  op_manual <- switch(
    op,
    d_separated = "d-separation",
    op
  )

  op_title <- tools::toTitleCase(gsub("_", " ", op_manual))
  ggplot(d, aes(n, median_ms, color = package)) +
    geom_line() +
    geom_point() +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(
      ~avg_degree,
      labeller = as_labeller(c("3" = "Sparser (avg degree 3)", "6" = "Denser (avg degree 6)"))
    ) +
    labs(
      x = "Number of nodes",
      y = "Median time (ms)",
      color = NULL,
      title = op_title
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
}
```

### Relational Queries

Direct neighbor lookups (`parents`, `children`) are $O(1)$ for `caugi`,
`igraph`, `bnlearn`, and Tetrad---they all maintain adjacency tables. `pgmpy` is
a thin wrapper over NetworkX's predecessor/successor lookups. `pcalg` and `ggm`
scan a row or column of the adjacency matrix on each call ($O(n)$), with `ggm`
paying noticeably more R-level overhead. `dagitty` is the outlier: it re-derives
neighbors from its string DSL representation each call and pays a
linear-in-edges cost.

```{r}
#| label: plot-parents

plot_op("parents")
```

```{r}
#| label: plot-children

plot_op("children")
```

As you can see from the plots, Tetrad is generally fastest, with pgmpy and caugi
close behind. The $O(1)$ packages stay flat as $n$ grows; pcalg also looks
nearly flat in practice because the per-call overhead dominates its $O(n)$ row
scan at these sizes. ggm and dagitty, in contrast, pay visibly for their design
choice to re-derive neighbors on each call. Note that the times are generally
very small (in milliseconds), and the differences between the front-runners may
not be that important in practice and likely reflect overhead differences
between Python, R, and Java.

### Ancestors and Descendants

These require transitive closure (BFS/DFS over the directed graph). Here
`caugi`'s CSR pays off---both forward and reverse adjacency are precomputed, so
traversal is tight pointer-chasing in Rust.

```{r}
#| label: plot-ancestors

plot_op("ancestors")
```

```{r}
#| label: plot-descendants

plot_op("descendants")
```

`caugi` is the most consistent here: sub-millisecond timings on every fixture
for both ancestors and descendants. `igraph::subcomponent()` and `pgmpy`'s
NetworkX-backed traversal also stay fast across the grid. Tetrad and `pgmpy`
trade leads on ancestors, while Tetrad's descendants routine slows dramatically
on the dense $n = 1000$ fixture (into the hundreds of milliseconds). `bnlearn`
and `pcalg` are uneven, occasionally spiking on the denser graphs. `dagitty` is
again one to two orders of magnitude slower than the rest, paying its
DSL-reparsing cost on every call.

### Markov Blanket

The Markov blanket of a node $v$ is
`parents(v) ∪ children(v) ∪ parents(children(v))`. `caugi`, `bnlearn`, and
Tetrad have direct accessors backed by adjacency tables; `dagitty` re-derives it
from its DSL representation each call; `pgmpy` walks NetworkX edges. `pcalg` and
`ggm` do not expose a Markov-blanket helper and are omitted.

```{r}
#| label: plot-markov-blanket

plot_op("markov_blanket")
```

`caugi`, `pgmpy`, and `bnlearn` all resolve Markov blankets in microseconds
across the entire grid. Tetrad is comparably fast on small graphs but slows down
on the dense $n = 1000$ fixture. `dagitty` is again the outlier, into the
hundreds of milliseconds on the larger graphs because it re-derives the blanket
from its DSL representation each call.

### D-separation

For each fixture we compute a valid backdoor adjustment set with
`caugi::adjustment_set()` once, and pass that same conditioning set to every
package's d-separation routine. Tetrad uses the m-separation generalisation
(equivalent on DAGs). `pcalg::dsep()` follows Lauritzen's moralization-based
test on a `graphNEL`.

```{r}
#| label: plot-dsep

plot_op("d_separated")
```

`caugi`, Tetrad, and `pgmpy` cluster at the fast end, all sub-millisecond on
small graphs and within a few milliseconds at $n = 1000$. `dagitty` is roughly
two orders of magnitude slower, while `bnlearn` and `pcalg` pay the cost of
building a moralized graph on each call: `bnlearn` reaches tens of milliseconds
on the larger fixtures and `pcalg` reaches the hundreds, with `pcalg` slowest
overall.

### Subgraph Extraction

Subgraph extraction is where we explicitly test graph **building** performance.
For `caugi` we time `subgraph(cg, nodes) |> build()` together so the CSR rebuild
cost is captured. `igraph`, `bnlearn`, `pgmpy`, and Tetrad all return
adjacency-backed subgraphs without an analogous index rebuild step. `pcalg`
exposes no native subgraph constructor (users typically call `graph::subGraph()`
on the underlying `graphNEL`), so it is omitted here.

```{r}
#| label: plot-subgraph

plot_op("subgraph")
```

`caugi` and `igraph` finish in well under a millisecond even at $n = 1000$, with
`caugi` slightly ahead despite paying for a CSR rebuild on every call. Tetrad
sits in the single-digit-millisecond range at $n = 1000$. `bnlearn` and `pgmpy` are roughly two orders
of magnitude slower on the largest fixtures, copying nodes and edges into a
fresh object each call. The takeaway is that even though `caugi` graphs are
nominally expensive to construct, the rebuild after `subgraph()` is fast enough
to be competitive with adjacency-list constructors that do not rebuild any
indexes at all.

## Summary

Across the seven operations, `caugi` either leads or sits in the leading group
on R-side competitors:

- It matches or beats `igraph` and `bnlearn` on the $O(1)$ adjacency lookups
  (`parents`, `children`, Markov blanket).
- It is the fastest R package by a wide margin on transitive-closure operations
  (`ancestors`, `descendants`, d-separation), where the CSR's precomputed
  forward and reverse adjacency pays off.
- Even on `subgraph()`, where the CSR has to be rebuilt, performance is
  competitive with packages that do not rebuild any index.

Tetrad (compiled Java) and `pgmpy` (NetworkX-backed) consistently beat `caugi`
on the simplest lookups by a small constant factor, which is unsurprising given
the R-level dispatch overhead `caugi` pays on every call. Where Tetrad slows
down---descendants on dense large graphs, dense Markov blankets---`caugi` keeps
its lead.

The headline still holds: by frontloading work into the build step and querying
a CSR-backed structure, `caugi` keeps query times effectively constant in graph
size for the operations users hit most often. The price is paid up front, in the
rebuild step that runs lazily on the first query after a mutation.

## Reproducing these benchmarks

The harness lives in
[`tools/benchmark/`](https://github.com/frederikfabriciusbjerre/caugi/tree/main/tools/benchmark)
and uses [Task](https://taskfile.dev) to orchestrate the cross-language runners.
To reproduce just the R numbers (no Python or Java toolchains needed):

```sh
cd tools/benchmark
task fixtures && task r
```

`task r` prints a per-package, per-operation summary to the console and writes
the raw timings to `results/r.csv`. See `tools/benchmark/README.md` for the full
pipeline (`task all`) and per- language details.

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
