---
title: "Visualizing Causal Graphs with caugi"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Visualizing Causal Graphs with caugi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

```{r setup}
library(caugi)
```

The `caugi` package provides a flexible plotting system built on grid graphics
for visualizing causal graphs. This vignette demonstrates how to create plots
with different layout algorithms and customize their appearance.

## Basic Plotting

The simplest way to visualize a `caugi` graph is with the `plot()` function:

```{r basic-plot}
# Create a simple DAG
cg <- caugi(
  A %-->% B + C,
  B %-->% D,
  C %-->% D,
  class = "DAG"
)

# Plot with default settings
plot(cg)
```

By default, `plot()` automatically selects the best layout algorithm based on
your graph type. For graphs with only directed edges, it uses the Sugiyama
hierarchical layout. For graphs with other edge types, it uses the
Fruchterman-Reingold force-directed layout.

## Layout Algorithms

The `caugi` package provides four layout algorithms, each optimized for
different use cases.

### Sugiyama (Hierarchical Layout)

The Sugiyama layout [@sugiyama1981] is ideal for directed acyclic graphs
(DAGs). It arranges nodes in layers to emphasize hierarchical structure
and causal flow from top to bottom, minimizing edge crossings.

```{r sugiyama-layout}
# Create a more complex DAG
dag <- caugi(
  X1 %-->% M1 + M2,
  X2 %-->% M2 + M3,
  M1 %-->% Y,
  M2 %-->% Y,
  M3 %-->% Y,
  class = "DAG"
)

# Use Sugiyama layout explicitly
plot(dag, layout = "sugiyama", main = "Sugiyama")
```

**Best for:** DAGs, causal models, hierarchical structures

**Limitations:** Only works with directed edges

### Fruchterman-Reingold (Spring-Electrical)

The Fruchterman-Reingold layout [@fruchterman1991] uses a physical simulation
where edges act as springs and nodes repel each other like charged particles.
It produces organic, symmetric layouts with relatively uniform edge lengths.

```{r fruchterman-reingold}
# Create a graph with bidirected edges (ADMG)
admg <- caugi(
  A %-->% C,
  B %-->% C,
  A %<->% B, # Bidirected edge (latent confounder)
  class = "ADMG"
)

# Fruchterman-Reingold handles all edge types
plot(admg, layout = "fruchterman-reingold", main = "Fruchterman-Reingold")
```

**Best for:** General-purpose visualization, graphs with mixed edge types

**Advantages:** Fast, works with all edge types, produces balanced layouts

### Kamada-Kawai (Stress Minimization)

The Kamada-Kawai layout [@kamada1989] minimizes "stress" by making Euclidean
distances in the plot proportional to graph-theoretic distances. This produces
high-quality layouts that better preserve the global structure compared to
Fruchterman-Reingold.

```{r kamada-kawai}
# Create an undirected graph
ug <- caugi(
  A %---% B,
  B %---% C + D,
  C %---% D,
  class = "UG"
)

# Kamada-Kawai for publication-quality visualization
plot(ug, layout = "kamada-kawai", main = "Kamada-Kawai")
```

**Best for:** Publication-quality figures, when accurate distance
representation matters

**Advantages:** Better global structure preservation

### Bipartite Layout

The bipartite layout is designed for graphs with a clear two-group structure,
such as treatment/outcome or exposure/response relationships. It arranges
nodes in two parallel lines (rows or columns).

```{r bipartite-layout}
# Create a bipartite graph: treatments -> outcomes
bipartite_graph <- caugi(
  Treatment_A %-->% Outcome_1 + Outcome_2 + Outcome_3,
  Treatment_B %-->% Outcome_1 + Outcome_2,
  Treatment_C %-->% Outcome_2 + Outcome_3,
  class = "DAG"
)

# Horizontal rows (treatments on top, outcomes on bottom)
plot(
  bipartite_graph,
  layout = "bipartite",
  orientation = "rows"
)

# Vertical columns (treatments on right, outcomes on left)
plot(
  bipartite_graph,
  layout = "bipartite",
  orientation = "columns"
)
```

The bipartite layout automatically detects which nodes should be in which
partition based on incoming edges. Nodes with no incoming edges are placed
in one group, while nodes with incoming edges are placed in the other. You
can also specify the partition explicitly:

```{r bipartite-explicit}
# Create bipartite layout with explicit partition
partition <- c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
plot(
  bipartite_graph,
  layout = caugi_layout_bipartite,
  partition = partition,
  orientation = "rows"
)
```

**Best for:** Treatment-outcome structures, exposure-response models,
bipartite causal relationships

**Advantages:** Clear visual separation, emphasizes directed relationships
between groups

### Comparing Layouts

You can compute and examine layout coordinates directly using `caugi_layout()`:

```{r compare-layouts}
# Compute layouts
layout_sug <- caugi_layout(dag, method = "sugiyama")
layout_fr <- caugi_layout(dag, method = "fruchterman-reingold")
layout_kk <- caugi_layout(dag, method = "kamada-kawai")

# Examine coordinates
head(layout_sug)
```

## Customizing Plots

The `plot()` function provides extensive customization options for nodes,
edges, and labels.

### Node Styling

Customize node appearance with the `node_style` parameter:

```{r node-styling}
# Customize node colors and sizes
plot(
  cg,
  node_style = list(
    fill = "lightblue", # Fill color
    col = "darkblue", # Border color
    lwd = 2, # Border width
    padding = 4, # Text padding (mm)
    size = 1.2 # Size multiplier
  )
)
```

Available node style parameters:

- **Appearance (passed to `gpar()`)**: `fill`, `col`, `lwd`, `lty`, `alpha`
- **Geometry**: `padding` (text padding in mm), `size` (node size multiplier)

### Edge Styling

Customize edge appearance with the `edge_style` parameter. You can set global
options or customize each edge type separately:

```{r edge-styling-global}
# Global edge styling
plot(
  dag,
  edge_style = list(
    col = "darkgray", # Edge color
    lwd = 1.5, # Edge width
    arrow_size = 4 # Arrow size (mm)
  )
)
```

```{r edge-styling-per-type}
# Per-type edge styling for ADMG
plot(
  admg,
  layout = "fruchterman-reingold",
  edge_style = list(
    directed = list(col = "blue", lwd = 2),
    bidirected = list(col = "red", lwd = 2, lty = "dashed")
  )
)
```

Available edge style parameters:

- **Appearance (passed to `gpar()`)**: `col`, `lwd`, `lty`, `alpha`, `fill`
- **Geometry**: `arrow_size` (arrow length in mm)
- **Per-type options**: `directed`, `undirected`, `bidirected`, `partial`

### Label Styling

Customize node labels with the `label_style` parameter:

```{r label-styling}
# Customize label appearance
plot(
  cg,
  main = "Customized Labels",
  label_style = list(
    col = "white", # Text color
    fontsize = 12, # Font size
    fontface = "bold", # Font face
    fontfamily = "sans" # Font family
  ),
  node_style = list(
    fill = "navy" # Dark background for white text
  )
)
```

Available label style parameters (passed to `gpar()`):

- `col`, `fontsize`, `fontface`, `fontfamily`, `cex`

### Combining Customizations

Put it all together for a fully customized plot:

```{r full-customization}
# Create a publication-ready plot
plot(
  dag,
  layout = "sugiyama",
  node_style = list(
    fill = "#E8F4F8",
    col = "#2C5F77",
    lwd = 1.5,
    padding = 3,
    size = 1.1
  ),
  edge_style = list(
    col = "#555555",
    lwd = 1.2,
    arrow_size = 3.5
  ),
  label_style = list(
    col = "#1A1A1A",
    fontsize = 11,
    fontface = "bold"
  )
)
```

## Working with Different Graph Types

The plotting system works seamlessly with all supported graph types.

### Partially Directed Acyclic Graphs (PDAGs)

```{r pdag-plot}
# Create a PDAG with both directed and undirected edges
pdag <- caugi(
  A %-->% B,
  B %---% C, # Undirected edge
  C %-->% D,
  class = "PDAG"
)

plot(
  pdag,
  edge_style = list(
    directed = list(col = "blue"),
    undirected = list(col = "gray", lwd = 2)
  )
)
```

### Acyclic Directed Mixed Graphs (ADMGs)

```{r admg-plot}
# Create an ADMG with confounding
complex_admg <- caugi(
  X %-->% M1 + M2,
  M1 %-->% Y,
  M2 %-->% Y,
  M1 %<->% M2, # Latent confounder between mediators
  class = "ADMG"
)

plot(
  complex_admg,
  layout = "kamada-kawai",
  node_style = list(fill = "lavender"),
  edge_style = list(
    directed = list(col = "black", lwd = 1.5),
    bidirected = list(col = "red", lwd = 1.5, lty = "dashed", arrow_size = 3)
  )
)
```

### Undirected Graphs (UGs)

```{r ug-plot}
# Create a Markov network
markov <- caugi(
  A %---% B + C,
  B %---% D,
  C %---% D + E,
  D %---% E,
  class = "UG"
)

plot(
  markov,
  layout = "fruchterman-reingold",
  node_style = list(
    fill = "lightyellow",
    col = "orange",
    lwd = 2
  ),
  edge_style = list(col = "orange")
)
```

## Plot Composition

The `caugi` package provides intuitive operators for composing multiple plots
into complex layouts, similar to the patchwork package.

### Basic Composition

Use `+` or `|` for horizontal arrangement and `/` for vertical stacking:

```{r composition-basic}
# Create two different graphs
g1 <- caugi(
  A %-->% B,
  B %-->% C,
  class = "DAG"
)

g2 <- caugi(
  X %-->% Y,
  Y %-->% Z,
  X %-->% Z,
  class = "DAG"
)

# Create plots
p1 <- plot(g1, main = "Graph 1")
p2 <- plot(g2, main = "Graph 2")

# Horizontal composition (side-by-side)
p1 + p2
```

The `|` operator is an alias for `+`:

```{r composition-pipe}
# Equivalent to p1 + p2
p1 | p2
```

For vertical stacking, use the `/` operator:

```{r composition-vertical}
# Stack plots vertically
p1 / p2
```

### Nested Compositions

Compositions can be nested to create complex multi-plot layouts:

```{r composition-nested}
# Create a third graph
g3 <- caugi(
  M1 %-->% M2,
  M2 %-->% M3,
  class = "DAG"
)

p3 <- plot(g3, main = "Graph 3")

# Complex layout: two plots on top, one below
(p1 + p2) / p3
```

You can mix operators freely:

```{r composition-mixed}
# Three plots in various arrangements
(p1 + p2) / (p3 + p1)
```

### Configuring Spacing

The spacing between composed plots is controlled globally via `caugi_options()`:

```{r composition-spacing}
# Set custom spacing
caugi_options(plot = list(spacing = grid::unit(2, "lines")))

# Plots will now have more space between them
p1 + p2

# Reset to default
caugi_options(plot = list(spacing = grid::unit(1, "lines")))
```

## Global Plot Options

The `caugi_options()` function allows you to set global defaults for plot
appearance, which can be overridden on a per-plot basis.

### Setting Default Styles

```{r global-options}
# Configure global defaults
caugi_options(plot = list(
  node_style = list(fill = "lightblue", padding = 3),
  edge_style = list(arrow_size = 4, fill = "darkgray"),
  title_style = list(col = "blue", fontsize = 16)
))

# This plot uses the global defaults
plot(cg, main = "Using Global Defaults")

# Reset to package defaults
caugi_options(caugi_default_options())
```

### Per-Plot Overrides

Global options serve as defaults that can be overridden:

```{r override-options}
# Set global node color
caugi_options(plot = list(
  node_style = list(fill = "lightblue")
))

# Override for this specific plot
plot(cg,
  main = "Custom Colors",
  node_style = list(fill = "pink")
)

# Reset
caugi_options(caugi_default_options())
```

### Available Options

The following options can be configured under `plot`:

- **`spacing`**: A `grid::unit()` controlling space between composed plots
- **`node_style`**: List with `fill`, `padding`, and `size`
- **`edge_style`**: List with `arrow_size` and `fill`
- **`label_style`**: List of text parameters (see `grid::gpar()`)
- **`title_style`**: List with `col`, `fontface`, and `fontsize`

```{r query-options}
# View all current options
caugi_options()

# Query specific option
caugi_options("plot")
```

## Advanced Usage

### Manual Layouts

You can compute layouts separately and reuse them:

```{r manual-layout}
# Compute layout once
coords <- caugi_layout(dag, method = "sugiyama")

# The layout can be used for analysis or custom plotting
print(coords)

# Plot uses the same layout, calling caugi_layout internally
plot(dag, layout = "sugiyama")
```

### Integration with Grid Graphics

`caugi` plots are built on grid graphics, and provide access to the underlying
grid `grob` object in the `@grob` slot of the plot output. This allows for
further customization using grid functions.

```{r grid-integration}
# Create a plot
p <- plot(cg)

# The grob slot is a grid graphics object
class(p@grob)

# You can manipulate it with grid functions
library(grid)

# Draw the plot rotated by 30 degrees
pushViewport(viewport(angle = 30))
grid.draw(p@grob)
popViewport()
```

The composition operators work by manipulating these grid grobs, creating
flexible and performant multi-plot layouts without requiring external packages.

## References

