---
title: "Visualizing Causal Graphs with caugi"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Visualizing Causal Graphs with caugi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

```{r setup}
library(caugi)
```

The `caugi` package provides a flexible plotting system built on grid graphics
for visualizing causal graphs. This vignette demonstrates how to create plots
with different layout algorithms and customize their appearance.

## Basic Plotting

The simplest way to visualize a `caugi` graph is with the `plot()` function:

```{r basic-plot}
# Create a simple DAG
cg <- caugi(
  A %-->% B + C,
  B %-->% D,
  C %-->% D,
  class = "DAG"
)

# Plot with default settings
plot(cg)
```

By default, `plot()` automatically selects the best layout algorithm based on
your graph type. For graphs with only directed edges, it uses the Sugiyama
hierarchical layout. For graphs with other edge types, it uses the
Fruchterman-Reingold force-directed layout.

## Layout Algorithms

The `caugi` package provides four layout algorithms, each optimized for
different use cases.

### Sugiyama (Hierarchical Layout)

The Sugiyama layout [@sugiyama1981] is ideal for directed acyclic graphs
(DAGs). It arranges nodes in layers to emphasize hierarchical structure
and causal flow from top to bottom, minimizing edge crossings.

```{r sugiyama-layout}
# Create a more complex DAG
dag <- caugi(
  X1 %-->% M1 + M2,
  X2 %-->% M2 + M3,
  M1 %-->% Y,
  M2 %-->% Y,
  M3 %-->% Y,
  class = "DAG"
)

# Use Sugiyama layout explicitly
plot(dag, layout = "sugiyama", main = "Sugiyama")
```

**Best for:** DAGs, causal models, hierarchical structures

**Limitations:** Only works with directed edges

### Fruchterman-Reingold (Spring-Electrical)

The Fruchterman-Reingold layout [@fruchterman1991] uses a physical simulation
where edges act as springs and nodes repel each other like charged particles.
It produces organic, symmetric layouts with relatively uniform edge lengths.

```{r fruchterman-reingold}
# Create a graph with bidirected edges (ADMG)
admg <- caugi(
  A %-->% C,
  B %-->% C,
  A %<->% B, # Bidirected edge (latent confounder)
  class = "ADMG"
)

# Fruchterman-Reingold handles all edge types
plot(admg, layout = "fruchterman-reingold", main = "Fruchterman-Reingold")
```

**Best for:** General-purpose visualization, graphs with mixed edge types

**Advantages:** Fast, works with all edge types, produces balanced layouts

### Kamada-Kawai (Stress Minimization)

The Kamada-Kawai layout [@kamada1989] minimizes "stress" by making Euclidean
distances in the plot proportional to graph-theoretic distances. This produces
high-quality layouts that better preserve the global structure compared to
Fruchterman-Reingold.

```{r kamada-kawai}
# Create an undirected graph
ug <- caugi(
  A %---% B,
  B %---% C + D,
  C %---% D,
  class = "UG"
)

plot(ug, layout = "kamada-kawai", main = "Kamada-Kawai")
```

**Best for:** Publication-quality figures, when accurate distance
representation matters

**Advantages:** Better global structure preservation

### Bipartite Layout

The bipartite layout is designed for graphs with a clear two-group structure,
such as treatment/outcome or exposure/response relationships. It arranges
nodes in two parallel lines (rows or columns).

Here's an example bipartite causal graph with treatments and outcomes:

```{r bipartite-layout}
bipartite_graph <- caugi(
  Treatment_A %-->% Outcome_1 + Outcome_2 + Outcome_3,
  Treatment_B %-->% Outcome_1 + Outcome_2,
  Treatment_C %-->% Outcome_2 + Outcome_3,
  class = "DAG"
)
```

Horizontal rows (treatments on top, outcomes on bottom)

```{r bipartite-basic}
#| fig-width: 5
plot(
  bipartite_graph,
  layout = "bipartite",
  orientation = "rows"
)
```

Vertical columns (treatments on left, outcomes on right)

```{r bipartite-vertical}
#| fig-height: 4
plot(
  bipartite_graph,
  layout = "bipartite",
  orientation = "columns"
)
```

The bipartite layout automatically detects which nodes should be in which
partition based on incoming edges. Nodes with no incoming edges are placed
in one group, while nodes with incoming edges are placed in the other. You
can also specify the partition explicitly:

```{r bipartite-explicit}
#| fig-width: 5
partition <- c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
plot(
  bipartite_graph,
  layout = caugi_layout_bipartite,
  partition = partition,
  orientation = "rows"
)
```

**Best for:** Treatment-outcome structures, exposure-response models,
bipartite causal relationships

**Advantages:** Clear visual separation, emphasizes directed relationships
between groups

### Tiered Layouts

For graphs with more than two hierarchical levels, the tiered layout places
nodes in multiple parallel tiers. This is ideal for visualizing causal
structures with clear stages, such as exposures → mediators → outcomes.

First, create a simple three-tier causal graph:

```{r tiered-basic}
cg_tiered <- caugi(
  X1 %-->% M1 + M2,
  X2 %-->% M1 + M2,
  M1 %-->% Y,
  M2 %-->% Y
)
```

We can define tiers using a named list:

```{r tiered-named-list}
#| fig-height: 4
tiers <- list(
  exposures = c("X1", "X2"),
  mediators = c("M1", "M2"),
  outcome = "Y"
)

plot(cg_tiered, layout = "tiered", tiers = tiers, orientation = "rows")
```

The tiered layout supports three input formats:

- named lists,
- named numeric vectors, and
- `data.frame`s.


Named lists is the most intuitive format, and what 
we have already shown. But you can also use a named numeric vector:

```{r tiered-vector}
tiers_vector <- c(X1 = 1, X2 = 1, M1 = 2, M2 = 2, Y = 3)
plot(cg_tiered, layout = "tiered", tiers = tiers_vector, orientation = "columns")
```

Finally, you can use a `data.frame` to specify tiers directly.

```{r tiered-dataframe}
tiers_df <- data.frame(
  name = c("X1", "X2", "M1", "M2", "Y"),
  tier = c(1, 1, 2, 2, 3)
)

layout_df <- caugi_layout_tiered(cg_tiered, tiers_df, orientation = "rows")

plot(cg_tiered, layout = layout_df)
```

**Best for:** Multi-stage causal processes, mediation analysis, temporal
sequences, hierarchical structures

**Advantages:** Clear stage separation, flexible tier assignment, supports 2+
tiers, multiple input formats

### Comparing Layouts

You can compute and examine layout coordinates directly using `caugi_layout()`:

```{r compare-layouts}
layout_sug <- caugi_layout(dag, method = "sugiyama")
layout_fr <- caugi_layout(dag, method = "fruchterman-reingold")
layout_kk <- caugi_layout(dag, method = "kamada-kawai")

# Examine coordinates
head(layout_sug)
```

## Customizing Plots

The `plot()` function provides extensive customization options for nodes,
edges, and labels.

### Node Styling

You can customize the appearance of nodes using the `node_style` parameter.
Styles may be applied globally (to all nodes) or locally (to specific nodes).

Apply the same style to all nodes:

```{r node-styling-global}
plot(
  cg,
  node_style = list(
    fill = "lightblue", # Fill color
    col = "darkblue", # Border color
    lwd = 2, # Border width
    padding = 4, # Text padding (mm)
    size = 1.2 # Size multiplier
  )
)
```

Customize styles for individual nodes using the `by_node` option:

```{r node-styling-locally}
plot(
  cg,
  node_style = list(
    by_node = list(
      A = list(fill = "red", col = "blue", lwd = 2),
      B = list(padding = "2")
    )
  )
)
```

Available node style parameters:

- **Appearance (passed to `gpar()`)**: `fill`, `col`, `lwd`, `lty`, `alpha`
- **Geometry**: `padding` (text padding in mm), `size` (node size multiplier)

### Edge Styling

You can customize edge appearance using the `edge_style` parameter.
Styles can be applied globally, by edge type, by source node, or to individual edges.

Apply the same styling to all edges in the graph:

```{r edge-styling-global}
plot(
  dag,
  edge_style = list(
    col = "darkgray", # Edge color
    lwd = 1.5, # Edge width
    arrow_size = 4 # Arrow size (mm)
  )
)
```

Customize different edge types (e.g., directed vs. bidirected edges):

```{r edge-styling-per-type}
plot(
  admg,
  layout = "fruchterman-reingold",
  edge_style = list(
    directed = list(col = "blue", lwd = 2),
    bidirected = list(col = "red", lwd = 2, lty = "dashed")
  )
)
```

Apply styling to all edges from a given node:

```{r edge-styling-per-node}
plot(
  admg,
  layout = "fruchterman-reingold",
  edge_style = list(
    by_edge = list(
      A = list(col = "green", lwd = 2)
    )
  )
)
```

Target an individual edge between two nodes:

```{r edge-styling-per-specific-edge}
plot(
  admg,
  layout = "fruchterman-reingold",
  edge_style = list(
    by_edge = list(
      A = list(
        B = list(col = "orange", lwd = 3)
      )
    )
  )
)
```

The style precedence is as follows (highest to lowest):

1. Specific edge (`by_edge` with both from and to nodes)
2. All edges from a node (`by_edge` with only from node)
3. Per-type edge styles (`directed`, `undirected`, etc.)
4. Global edge styles

The example below combines global, per-type, per-node, and per-edge styling in a single plot.
More specific styles override more general ones according to the precedence rules above.

```{r edge-styling-combined}
plot(
  admg,
  layout = "fruchterman-reingold",
  edge_style = list(
    # Global defaults
    col = "gray80",
    lwd = 1,

    # Per-type styling
    directed = list(col = "blue"),
    bidirected = list(col = "red", lty = "dashed"),

    # All edges from node A
    by_edge = list(
      A = list(
        col = "green",
        lwd = 2,

        # Specific edge A -> B
        B = list(
          col = "orange",
          lwd = 3
        )
      )
    )
  )
)
```

Available edge style parameters:

- **Appearance (passed to `gpar()`)**: `col`, `lwd`, `lty`, `alpha`, `fill`
- **Geometry**: `arrow_size` (arrow length in mm), `circle_size` (radius of endpoint circles for partial edges in mm)
- **Per-type options**: `directed`, `undirected`, `bidirected`, `partial`

#### Partial Edges

Partial edges (`o->` and `o-o`) are rendered with circles at their endpoints to
indicate uncertainty about edge orientation. These edges appear in PAGs
(Partial Ancestral Graphs). You can customize the circle size:

```{r partial-edges}
g <- caugi(
  A %o->% B,
  B %-->% C,
  C %o-o% D,
  class = "UNKNOWN"
)

plot(
  g,
  edge_style = list(
    partial = list(
      col = "purple",
      lwd = 2,
      circle_size = 2.5 # Larger circles (default is 1.5)
    )
  )
)
```

### Label Styling

Customize node labels with the `label_style` parameter:

```{r label-styling}
plot(
  cg,
  main = "Customized Labels",
  label_style = list(
    col = "white", # Text color
    fontsize = 12, # Font size
    fontface = "bold", # Font face
    fontfamily = "sans" # Font family
  ),
  node_style = list(
    fill = "navy" # Dark background for white text
  )
)
```

Available label style parameters (passed to `gpar()`):

- `col`, `fontsize`, `fontface`, `fontfamily`, `cex`

### Styling Tiered Layouts

By default, tiered layouts are plotted with boxes around each tier
and labels indicating the tier names (if provided).

```{r tiered-boxes-basic}
plot(cg_tiered, tiers = tiers)
```

But you can customize the appearance of tier boxes using the `tier_style` parameter.
Here, for isntance, we specify different fill colors for each tier using a vector.

```{r tiered-boxes-vector}
plot(
  cg_tiered,
  tiers = tiers,
  tier_style = list(
    fill = c("lightblue", "lightgreen", "lightyellow"),
    col = "gray50",
    lty = 2,
    alpha = 0.3
  )
)
```

For more granular control, you can specify styles for individual tiers. 
Here, we customize the "exposures" and "outcome" tiers specifically:

```{r tiered-boxes-by-tier}
plot(
  cg_tiered,
  tiers = tiers,
  tier_style = list(
    fill = "gray95",
    col = "gray60",
    alpha = 0.2,
    by_tier = list(
      exposures = list(
        fill = "lightblue",
        col = "blue",
        lwd = 2
      ),
      outcome = list(
        fill = "lightyellow",
        col = "orange",
        lwd = 3,
        lty = 1
      )
    )
  )
)
```

Labels for the tiers can also be customized. 
Here, for example, we change the font size and color of the tier labels:

```{r tiered-boxes-labels}
plot(
  cg_tiered,
  tiers = tiers, # Named list: exposures, mediators, outcome
  tier_style = list(
    fill = c("lightblue", "lightgreen", "lightyellow"),
    label_style = list(
      fontsize = 11,
      fontface = "bold",
      col = "gray20"
    )
  )
)
```

You can also provide custom labels for each tier instead of using the
names from the tiers object

```{r tiered-boxes-custom-labels}
#| fig-width: 5
plot(
  cg_tiered,
  tiers = tiers,
  tier_style = list(
    fill = "gray95",
    labels = c("Exposure Variables", "Mediating Variables", "Outcome Variable")
  )
)
```

If you don't want any boxes or labels around the tiers, you can disable them:

```{r tiered-boxes-none}
plot(
  cg_tiered,
  tiers = tiers,
  tier_style = list(boxes = FALSE, labels = FALSE)
)
```

## Working with Different Graph Types

The plotting system works with all graph types supported by `caugi`.

### Partially Directed Acyclic Graphs (PDAGs)

First, let's create a PDAG with both directed and undirected edges:

```{r pdag-plot}
pdag <- caugi(
  A %-->% B,
  B %---% C, # Undirected edge
  C %-->% D,
  class = "PDAG"
)

plot(
  pdag,
  edge_style = list(
    directed = list(col = "blue"),
    undirected = list(col = "gray", lwd = 2)
  )
)
```

### Acyclic Directed Mixed Graphs (ADMGs)

Here's an example of an ADMG with directed and bidirected edges:

```{r admg-plot}
complex_admg <- caugi(
  X %-->% M1 + M2,
  M1 %-->% Y,
  M2 %-->% Y,
  M1 %<->% M2, # Latent confounder between mediators
  class = "ADMG"
)

plot(
  complex_admg,
  layout = "kamada-kawai",
  node_style = list(fill = "lavender"),
  edge_style = list(
    directed = list(col = "black", lwd = 1.5),
    bidirected = list(col = "red", lwd = 1.5, lty = "dashed", arrow_size = 3)
  )
)
```

### Undirected Graphs (UGs)

We also support undirected graphs. Here's a Markov random field example:

```{r ug-plot}
markov <- caugi(
  A %---% B + C,
  B %---% D,
  C %---% D + E,
  D %---% E,
  class = "UG"
)

plot(
  markov,
  layout = "fruchterman-reingold",
  node_style = list(
    fill = "lightyellow",
    col = "orange",
    lwd = 2
  ),
  edge_style = list(col = "orange")
)
```

## Plot Composition

The `caugi` package provides intuitive operators for composing multiple plots
into complex layouts, similar to the patchwork package.

### Basic Composition

Use `+` or `|` for horizontal arrangement and `/` for vertical stacking:

```{r composition-basic}
# Create two different graphs
g1 <- caugi(
  A %-->% B,
  B %-->% C,
  class = "DAG"
)

g2 <- caugi(
  X %-->% Y,
  Y %-->% Z,
  X %-->% Z,
  class = "DAG"
)

# Create plots
p1 <- plot(g1, main = "Graph 1")
p2 <- plot(g2, main = "Graph 2")

# Horizontal composition (side-by-side)
p1 + p2
```

The `|` operator is an alias for `+`:

```{r composition-pipe}
# Equivalent to p1 + p2
p1 | p2
```

For vertical stacking, use the `/` operator:

```{r composition-vertical}
#| fig-height: 5
p1 / p2
```

### Nested Compositions

Compositions can be nested to create complex multi-plot layouts:

```{r composition-nested}
#| fig-height: 5
g3 <- caugi(
  M1 %-->% M2,
  M2 %-->% M3,
  class = "DAG"
)

p3 <- plot(g3, main = "Graph 3")

# Complex layout: two plots on top, one below
(p1 + p2) / p3
```

You can mix operators freely. Here's an example combining horizontal and vertical arrangements:

```{r composition-mixed}
#| fig-height: 5
(p1 + p2) / (p3 + p1)
```

### Configuring Spacing

The spacing between composed plots is controlled globally via `caugi_options()`:

```{r composition-spacing}
caugi_options(plot = list(spacing = grid::unit(2, "lines")))

p1 + p2
```

To reset the default, you can call
`caugi_default_options()`:

```{r reset-global-options}
caugi_options(caugi_default_options())
```

## Global Plot Options

The `caugi_options()` function allows you to set global defaults for plot
appearance, which can be overridden on a per-plot basis.

### Setting Default Styles

```{r global-options}
# Configure global defaults
caugi_options(plot = list(
  node_style = list(fill = "lightblue", padding = 3),
  edge_style = list(arrow_size = 4, fill = "darkgray"),
  title_style = list(col = "blue", fontsize = 16)
))

# This plot uses the global defaults
plot(cg, main = "Using Global Defaults")
```

### Per-Plot Overrides

Global options serve as defaults that can be overridden:

```{r override-options}
# Set global node color
caugi_options(plot = list(
  node_style = list(fill = "lightblue")
))

# Override for this specific plot
plot(cg,
  main = "Custom Colors",
  node_style = list(fill = "pink")
)

# Reset to defauls
caugi_options(caugi_default_options())
```

### Available Options

The following options can be configured under `plot`:

- **`spacing`**: A `grid::unit()` controlling space between composed plots
- **`node_style`**: List with `fill`, `padding`, and `size`
- **`edge_style`**: List with `arrow_size` and `fill`
- **`label_style`**: List of text parameters (see `grid::gpar()`)
- **`title_style`**: List with `col`, `fontface`, and `fontsize`

```{r query-options}
# View all current options
caugi_options()

# Query specific option
caugi_options("plot")
```

## Advanced Usage

### Manual Layouts

You can compute layouts separately and reuse them.
First, compute the layout coordinates:

```{r manual-layout}
coords <- caugi_layout(dag, method = "sugiyama")

# The layout can be used for analysis or custom plotting
print(coords)

# Plot uses the same layout, calling caugi_layout internally
plot(dag, layout = "sugiyama")
```

### Integration with Grid Graphics

`caugi` plots are built on grid graphics, and provide access to the underlying
grid `grob` object in the `@grob` slot of the plot output. This allows for
further customization using grid functions.

```{r grid-integration}
# Create a plot
p <- plot(cg)

# The grob slot is a grid graphics object
class(p@grob)

# You can manipulate it with grid functions
library(grid)

# Draw the plot rotated by 30 degrees
pushViewport(viewport(angle = 30))
grid.draw(p@grob)
popViewport()
```

The composition operators work by manipulating these grid grobs, creating
flexible and performant multi-plot layouts without requiring external packages.

## References

